\chapter{Phase 3: DOSBox and Hugo}

\section{Goal}

Have Hugo running on the Pi, displayed on the HDMI screen and playable with a USB keyboard. This is the last individual piece before tying everything together in Phase 4.

\section{Legal status of the game}

Hugo is \emph{abandonware}: commercial software whose support by the original publisher ceased decades ago, and whose passive distribution is broadly tolerated though not formally authorized. Sites like \texttt{myabandonware.com} and \texttt{abandonwaregames.net} host downloadable copies. The legal situation varies by jurisdiction, but home and educational use has not been the subject of litigation.

\section{Step 1: Get Hugo}

The series has several entries and the format varies:

\begin{itemize}
  \item \textbf{Hugo 1 / 2 / 3} (1990--1992) come as a ZIP that, once extracted, contains a ready-to-run \texttt{HUGO.EXE}.
  \item \textbf{Hugo 4 and later} are usually distributed as a CD-ROM ISO image with \texttt{INSTALL.EXE}, \texttt{INSTALLW.EXE} and \texttt{HUGO.EXE}. They require running the installer inside DOSBox the first time.
\end{itemize}

Once downloaded, transfer the file to the Pi:

\begin{shellcode}
# From the Mac, ZIP (Hugo 1/2/3):
scp ~/Downloads/Hugo*.zip beto@hugo-pbx.local:/tmp/

# From the Mac, ISO (Hugo 4 and other CD-ROM releases):
scp ~/Downloads/hugo4.iso \
  beto@hugo-pbx.local:/opt/hugo-phone/games/
\end{shellcode}

\subsection{Case A: ZIP with a loose \texttt{HUGO.EXE}}

\begin{shellcode}
sudo mkdir -p /opt/hugo-phone/games/HUGO
sudo unzip /tmp/Hugo*.zip -d /opt/hugo-phone/games/HUGO
sudo chown -R beto:beto /opt/hugo-phone/games/HUGO
ls /opt/hugo-phone/games/HUGO/
\end{shellcode}

You should see \texttt{HUGO.EXE} (or a similar file). Skip to Step 4.

\subsection{Case B: CD-ROM ISO (Hugo 4)}

The ISO stays as a file. DOSBox mounts it as drive \texttt{D:} with \texttt{imgmount}. The first time you must run the installer (\texttt{INSTALL.EXE}, not \texttt{INSTALLW.EXE}, which targets Windows) so it copies files to \texttt{C:\textbackslash{}HUGO\textbackslash{}} and configures audio.

\section{Step 2: Install DOSBox}

Classic DOSBox is in the official Raspberry Pi OS / Debian Trixie repos:

\begin{shellcode}
sudo apt update
sudo apt install -y dosbox xinit openbox xserver-xorg
which dosbox
\end{shellcode}

\begin{note}
The \texttt{dosbox-staging} package is not available in the standard Pi repositories at the time of writing. If it becomes available later, the options in \texttt{dosbox-hugo.conf} would need to be adapted to \emph{Staging}'s syntax.
\end{note}

\section{Step 3: Game installer (Case B / ISO only)}

Plug HDMI + a USB keyboard into the Pi and log in on the local TTY (not over SSH). DOSBox needs a display.

Launch the installer with the ISO mounted as \texttt{D:}:

\begin{shellcode}
startx /usr/bin/dosbox -- \
  -c "mount c /opt/hugo-phone/games" \
  -c "imgmount d /opt/hugo-phone/games/hugo4.iso -t iso" \
  -c "d:" \
  -c "INSTALL.EXE"
\end{shellcode}

Inside the installer:

\begin{itemize}
  \item \textbf{Destination}: \texttt{C:\textbackslash{}HUGO} (or whatever it defaults to).
  \item \textbf{Sound card}: Sound Blaster 16, base \texttt{220}, IRQ \texttt{7}, DMA \texttt{1}. Matches the \texttt{[sblaster]} section of \texttt{dosbox-hugo.conf}.
  \item Accept and exit the installer.
\end{itemize}

Verify the files were copied:

\begin{shellcode}
ls /opt/hugo-phone/games/HUGO/
\end{shellcode}

\section{Step 4: Verify DOSBox configuration}

The install script copied \texttt{dosbox-hugo.conf} to \texttt{/opt/hugo-phone/games/}. Verify:

\begin{shellcode}
head -30 /opt/hugo-phone/games/dosbox-hugo.conf
\end{shellcode}

If the game's executable is not exactly \texttt{HUGO.EXE} or lives in a different subfolder, edit the \texttt{[autoexec]} section:

\begin{shellcode}
nano /opt/hugo-phone/games/dosbox-hugo.conf
\end{shellcode}

Default section (Case B, Hugo 4 ISO with CD present):

\begin{plaincode}
[autoexec]
mount c /opt/hugo-phone/games
imgmount d /opt/hugo-phone/games/hugo4.iso -t iso
c:
cd HUGO
HUGO.EXE
exit
\end{plaincode}

For Case A (loose HUGO.EXE, no ISO), comment out the \texttt{imgmount} line.

\section{Step 5: Launch Hugo}

DOSBox from the Debian package uses SDL 1.2 and requires an X server (no direct KMSDRM support). The simplest path is \texttt{startx} + DOSBox inside:

\begin{shellcode}
startx /usr/bin/dosbox -conf /opt/hugo-phone/games/dosbox-hugo.conf
\end{shellcode}

This starts a minimal X11, runs DOSBox inside, and when DOSBox closes, X terminates and you're back at the TTY.

\begin{note}
\texttt{startx} only works from a local TTY (not over SSH). For remote testing you can use SSH with X forwarding (\texttt{ssh -Y}), but performance is much worse and is not recommended for real playability.
\end{note}

\section{Step 6: Auto-start the game}

Configure the Pi so that powering on takes you directly into Hugo, with no user intervention.

\subsection{Enable autologin on TTY1}

\begin{shellcode}
sudo raspi-config
\end{shellcode}

Navigate to: \textsf{System Options} $\to$ \textsf{Boot / Auto Login} $\to$ \textsf{Console Autologin}. Save and exit.

\subsection{Run DOSBox at login}

Edit the \texttt{.bash\_profile} of the \texttt{beto} user:

\begin{shellcode}
nano ~/.bash_profile
\end{shellcode}

Append at the end:

\begin{plaincode}
if [ -z "$DISPLAY" ] && [ "$(tty)" = "/dev/tty1" ]; then
  startx /usr/bin/dosbox \
    -conf /opt/hugo-phone/games/dosbox-hugo.conf
fi
\end{plaincode}

The condition prevents DOSBox from running if you log in over SSH or from a TTY other than 1.

After reboot, the sequence should be:

\begin{enumerate}
  \item Kernel boot
  \item Autologin as \texttt{beto} on \texttt{tty1}
  \item \texttt{startx} brings up X11
  \item DOSBox starts, runs \texttt{[autoexec]} and Hugo appears
\end{enumerate}

\section{Step 7: Playability test}

With a USB keyboard plugged into the Pi and Hugo loaded, test:

\begin{itemize}
  \item Arrow keys $\to$ character movement
  \item Space bar $\to$ action
  \item \texttt{Esc} $\to$ menu/exit
\end{itemize}

If Hugo feels excessively fast or slow, adjust the \texttt{cycles} line in \texttt{dosbox-hugo.conf}:

\begin{tomlcode}
[cpu]
cycles = fixed 8000   # Increase if too slow, decrease if too fast
\end{tomlcode}

The original Hugo required roughly a 386SX at 25\,MHz. Start with \texttt{cycles = fixed 8000} and calibrate.

\section{Step 8: Exit cleanly}

Two \texttt{Esc} presses bring up Hugo's menu. The \emph{Quit} option returns to the DOSBox prompt; the \texttt{exit} command in the \texttt{[autoexec]} block closes DOSBox automatically.

For an emergency exit (for example if the game hangs): \texttt{Ctrl}+\texttt{F9} is the \emph{kill switch} for DOSBox.

\section{Troubleshooting}

\subsection{\texttt{startx} can't find display or fails}

Confirm the X server packages are present:

\begin{shellcode}
sudo apt install -y xinit openbox xserver-xorg
\end{shellcode}

And that you're logged into a local TTY (not SSH). Check with \texttt{tty} (it should answer \texttt{/dev/tty1} or similar, not \texttt{/dev/pts/*}).

\subsection{Hugo hangs on the intro or the cutscenes}

DOSBox on ARM occasionally has problems decoding VGA animations. Pressing any key to skip the intro usually solves it. If it persists, lower the \emph{output} quality:

\begin{tomlcode}
[sdl]
output = surface
\end{tomlcode}

\subsection{The audio stutters}

Increase the mixer buffer:

\begin{tomlcode}
[mixer]
prebuffer = 50
blocksize = 2048
\end{tomlcode}

If stuttering remains, lower the quality: \texttt{rate = 22050}.

\subsection{The game asks for CD but the ISO is mounted}

Make sure \texttt{imgmount} didn't fail silently:

\begin{shellcode}
ls -la /opt/hugo-phone/games/hugo4.iso
\end{shellcode}

It must exist and be readable by the \texttt{beto} user. If not, fix permissions:

\begin{shellcode}
sudo chown beto:beto /opt/hugo-phone/games/hugo4.iso
\end{shellcode}

\subsection{Hugo starts but doesn't respond to the keyboard}

\begin{itemize}
  \item Verify that \texttt{core = dynamic} is present in the \texttt{[cpu]} section.
  \item Some titles require \texttt{cputype = 386} instead of \texttt{auto}.
  \item Make sure the DOSBox window has focus (with \texttt{autolock = true} the keyboard is captured on click).
\end{itemize}

\section{Phase wrap-up}

At this point the game runs with a USB keyboard. The next phase replaces that USB keyboard with a virtual one controlled from the phone.
